The compositions of high-pressure water-rich melts in the system MgO-SiO$_2$-H$_2$O

### 2015-07-07 MSH.tex
Update

\title{The compositions of high-pressure water-rich melts in the system MgO-SiO$_2$-H$_2$O}
%\tnotetext[mytitlenote]{Fully documented templates are available in the elsarticle package on \href{http://www.ctan.org/tex-archive/macros/latex/contrib/elsarticle}{CTAN}.}

@@

\begin{abstract}
High pressure multianvil experiments were conducted at 13 GPa between 1200 and 1900 $^{\circ}$C to investigate the liquidus in the system MgO-SiO$_2$-H$_2$O. Water-rich starting compositions were obtained by using platinic acid (H$_2$Pt(OH)$_6$) as a novel water source. The topology of the liquidus is tightly constrained at much higher water contents than previously possible. In the subsystem MgO-H$_2$O, the invariant point between brucite, periclase and liquid occurs at ca. 1210 $^{\circ}$C and 67 mol\% H$_2$O. Forsterite and clinoenstatite both melt congruently in the temperature range investigated, although the fo-cen and cen-stv cotectics migrate close to the fo-H$_2$O and cen-H$_2$O binaries at high water contents. The chondrodite liquidus field approaches the fo-H$_2$O binary at ca. 1200 $^{\circ}$C. Preliminary experiments on the SiO$_2$-H$_2$O binary reveal a pronounced melting point depression up to about 20 mol\% H$_2$O, then shallowing up to 50 mol\%, similar to behaviour previously observed at 2 GPa.
\end{abstract}

@@

\section{Introduction}
The subduction of hydrous materials has a strong influence on the formation of melts within the deep Earth. The dehydration of hydrous crustal minerals is thought to drive the melting of the mantle wedge that feeds arc volcanoes. Dehydration of the underlying lithospheric mantle may also enable embrittlement of the slab, allowing it to deform at significantly higher strain rates than would be possible if it were dry.

There has been some controversy over whether slabs transport water beyond about 300 km depth. Certainly most of the ca. 7 km thick mafic crust is probably too warm to sustain hydrous minerals. However, hydrated ultramafics can potentially carry water into the lower mantle. Supporting evidence for an ultradeep part of the water cycle exists in the form of hydrous ringwoodite in diamond, which apparently formed under cold conditions.

The compositions of melts formed by the breakdown of hydrous phases is important for a full understanding of the deep water system. The thermodynamics of melting in water-rich systems cannot be determined without good estimates of melt compositions. Furthermore, the thermal stability of hydrous phases cannot be predicted from their properties without also having an activity-composition model for the melt phase. More generally, the density of hydrous melts will be highly composition dependent, impacting on their ability to rise through the mantle. Finally, reactions between migrating melts and the surrounding solid will be controlled by the \emph{P-T-X} dependence on melt composition.

Although the phase relations in hydrous systems in the upper mantle are now remarkably well-constrained, constraining melt compositions has remained a major challenge in high-pressure experimental petrology. The primary problem is that hydrous melts are unquenchable, which means that their compositions must be estimated indirectly. Compositions of partial hydrous melts have been estimated using mass balance, diamond traps \citep{MSUP2007}, and the Mg:Si ratio has been estimated by defocusing the beam of an electron probe microanalyser (EPMA) \citep[e.g.][]{YII2004}. Unfortunately, all three of these techniques are accompanied by significant uncertainties, especially in terms of the water content. An alternative technique is to constrain the liquidus temperature at a given composition. However, solid hydrous compounds do not reach the water concentrations expected from hydrous melts in the vicinity of a subducting slab. Thus, studies either extrapolate high temperature liquidus curves to high water contents or add water in liquid form. Extrapolation is prone to large errors, especially when high temperatures facilitate water loss through capsule walls. Water masses can be accurately calculated when added to large volume chambers, but not to the small volumes of high pressure capsules. Thus, constraining the water contents of melts in hydrous systems remains a key problem in experimental petrology.

@@

A number of models have been presented to investigate the behaviour of concentrated hydrous melts. \cite{SS1985} made the assumption that H$_2$O, O$_{br}^{2-}$ and OH$_{tm}^-$ mix ideally, with a parameter $r$ representing the number of oxygen atoms in each formula unit which are available for protonation. For example, an Mg$_2$SiO$_4$ melt with $r=4$ and $K=\infty$ represents a continuum between an anhydrous melt and one made purely of monomers. If r is less then the number of oxygens per formula unit no distinction is made between isolated oligomers and partial depolymerisation of a silicate network; in other words, the energy required to form a terminal OH group is independent of local environment. Fixing $r$ places an implicit constraint on the maximum value of $K$ in water-rich compositions. For example, if $r=2$ in a hydrous forsteritic melt, then water contents exceeding 50\% require that $K<\infty$. In practise, the equilibrium constant $K$ has been shown to be dependent on temperature and composition. \cite{SK1995} obtained a good fit to isochemical data with an expression of the form
\begin{equation}
\Delta G^{\circ} = a + b\,T
\end{equation}

@@

In all of these models, the activity of component $x$ with the composition of the liquidus phase $X$ can be calculated from the melt model and independently from the entropy of melting of $X$. At equilibrium, these two activities must be equal:
\begin{equation}
RT \ln a_{x,L}^X = G_{X_l} - G_{X_{s}} = - \int_T^{T_{melt}} S_{X_{l}}-S_{X_s} \, dT
\end{equation}
where $T_{melt}$ is the temperature of melting of the liquidus phase under anhydrous conditions.

@@

\subsection{Subsystems}
\subsubsection{MgO-H$_2$O}
The dehydration of brucite at high pressure was previously studied by \cite{FIYKFO2005}. They reported that the stability of brucite reaches a maximum at about 9--10 GPa and 1200 $^{\circ}$C, decomposing at 1100--1150 $^{\circ}$C at 13 GPa. We observed large brucite crystals at 1200 $^{\circ}$C and 13 GPa; so assuming that errors from both studies are $\sim$ 50 $^{\circ}$C, decomposition takes place just above 1200 $^{\circ}$C. In the present study, the composition of the fluid coexisting with brucite and periclase is constrained to be 64--66 mol\% H$_2$O. The MgO content of the fluid increases gradually with increasing pressure, as a result of the extremely high melting temperature of periclase. The best experimental estimate of the melting point of MgO is currently that of \cite{ZF2008}, who extrapolated to the MgO endmember using melting phase relations in the system (Mg,Fe)O. Their melting curve is essentially identical to that obtained from large-scale molecular dynamics simulations \citep{CG1994}. Using these two studes, melting at 13 GPa is estimated to be $\sim 5373$ K. This value is higher than that from more recent ab-initio investigations \citep{Alfe2005,KKS2013}, and 2000 K higher than the melting temperature estimated experimentally by \cite{ZB1994}.

At 13 GPa, the entropy of melting of periclase is estimated to be $\sim$22 J/K/mol \cite{CG1994}. Assuming that the metastable entropy of melting at this pressure remains constant with temperature, the activity of periclase in the system can be determined and used to fit a melting model to the experimental data. Using the ideal OH$^-$ mixing model of \cite{SS1985} with no compositional dependence on the formation of OH groups, $G^{\circ}= -70000 - 15T$ provides a good fit to the data. However, this model yields a low activity of water in equilibrium with brucite and periclase ($a_{H_2O}\sim 0.34$) relative to that estimated using the thermodynamic data of \citep{HP2011} and the equation of state of water formulated by \cite{PS1995} ($0.52$).

To simultaneously fit the melting data and the activity of water in melts in equilibrium with periclase and brucite, $\partial^2 G_{ex} / \partial X_{H_2O}^2$ needs to be less positive at compositions more water-rich than brucite. This can be achieved by imposing a binary subregular model (W$_{MgO-H_2O}$=38000 J/mol, W$_{H_2O-MgO}$=0 J/mol). This model also has the effect of increasing the tendency for melt-fluid immiscibility at the water-rich end of the system. The second critical endpoint of the MgO-H$_2$O is probably around 13 GPa \citep{MSUP2007}, so large positive interaction enthalpies are justified in this region of the phase diagram. Alternatively, a regular solution model with MgO, Mg(OH)$_2$ and H$_2$O species in the melt can adequately fit the data, with parameters G$_{br,melt} = -41000. - 10*(T-(1210.+273.15))$, W$_{MgO-H_2O}=0$, W$_{MgO-Mg(OH)_2}=-10000$, W$_{Mg(OH)_2-H_2O}=-22000$ (all in J/mol). The resulting fits are shown in Figure \ref{fig:MH}. The large differences between the two model-types at high temperature is a result of the much larger entropy of mixing in the \cite{SS1985} models close to the anhydrous endmembers.

@@

One chamber from an experiment run at 1300 $^{\circ}$C revealed a single very small crystal of enstatite. Another run at  $^{\circ}$C revealed large crystals of hydroxychondrodite in the chamber with the lowest water contents. Small differences in Mg:Si ratio could explain the differences in solid phases between the separate chambers, so it is assumed that the enstatite+forsterite and chondrodite+forsterite cotectics migrate very close to the forsterite+H$_2$O binary at high water contents and approximately 1200 $^{\circ}$C.

@@

\subsection{Hydrous melting in high pressure subduction zone systems}
Subducting slabs are thought of transport significant quantities of mineral-bound water into the mantle transition zone. Much of this water is lost within the upper 200--300 km of the mantle through the breakdown of hydrous minerals such as antigorite and lawsonite. However, some mineral-bound water is thought to be preserved within a series of dense hydrous magnesium silicate (DHMS) phases. This observation was recently confirmed by the discovery of hydrous ringwoodite in diamond, which must have formed in the lower half of the mantle transition zone. The amount of water in the ringwoodite suggested that last equilibration occured under cold conditions probably associated with subduction. A key question, then, is how this material reached the surface sufficiently quickly to avoid retrogression to olivine, and how it remained cold during ascent to avoid the creation of hydrous melts. One effective way to allow melts to rapidly ascend to the surface is to channelise them. In turn, channelisation is promoted by reaction-infiltration and positive deformation-infiltration feedbacks.

The results of this study provide important constraints on the relationships between hydrous melt and the surrounding environment under the conditions at the top of the mantle transition zone. Within a single lithology, melts percolating upwards into hotter surroundings will consume small amounts of the surrounding solids, encouraging channelisation. At such high pressures, the change in fluid composition with temperature is gradual along the forsterite-enstatite and enstatite-stishovite cotectics, and the stishovite melting curve. Thus, it is unlikely that melt fraction will rapidly change in a single lithology.

A more profound effect occurs when hydrous fluids in equilibrium with peridotites infiltrate more Si-rich rocks. The enstatite-stishovite cotectic has a much greater $\partial T/\partial X_{H_2O}$ than the forsterite-enstatite cotectic at high temperatures. This implies that fluids entering silica-rich rocks will be corrosive, consuming enstatite and stishovite to reach equilibrium with their new surroundings. Despite being volumetrically minor components of subduction zone systems, silica-rich lithologies could play a dominant role in channelising hydrous fluids. The motion of these fluids could also significantly weaken the uppermost slab, effectively decoupling it from the overlying mantle wedge. Such effects promote trench parallel flow, encouraging slab deformation and rollback (trench migration relative to the surrounding mantle).

>>>>

\title{Hydrous transition zone melting  
%\title{High-pressure water-rich melting in the system MgO-SiO$_2$-H$_2$O: Insights into hydrous transition zone melting}
%\tnotetext[mytitlenote]{Fully documented templates are available in the elsarticle package on \href{http://www.ctan.org/tex-archive/macros/latex/contrib/elsarticle}{CTAN}.}

@@

\begin{abstract}

In this study, we conducted high pressure multianvil experiments at 13 GPa between 1200 and 1900 $^{\circ}$C to investigate the liquidus in the system MgO-SiO$_2$-H$_2$O. Water-rich starting compositions were obtained by using platinic acid (H$_2$Pt(OH)$_6$) as a novel water source. As MgO:SiO$_2$ ratios decrease, the lquidus curve develops an increasingly pronounced concave-down topology. At slab temperatures, the forsterite- and enstatite-water liquidii have similar water contents. Preliminary experiments on the SiO$_2$-H$_2$O binary reveal a pronounced melting point depression up to about 20 mol\% H$_2$O. This melting point depression exceeds estimates from published models, and requires either significant non-ideality between species in the melt, or distinct differences in the ease of protonation of bridging oxygens. We discuss the implications of this work on the migration of deep-seated slab fluids, and the partitioning of water between nominally anhydrous phases and melts.
\end{abstract}

@@

\section{Introduction}
Subduction zones are primary vectors for water transport into the deep Earth. This water is progressively lost as fluids and then melts as the surrounding mantle warms the slab. The most prominent surface expression of this melting is the presence of arc volcanoes about 100 km above the slab surface, but hydrous minerals within the crust should allow some water to be transported to 250 km, and peridotite hydration should allow water to penetrate into the lowermost mantle transition zone, providing it remains low enough in temperature. A startling confirmation of such deep water is the discovery of water-saturated ringwoodite in a kimberlitic diamond from Juina, Brazil.

Several key questions present themselves. When water-rich melts are created in subduction zones, what are their compositions? How do their compositions and resulting physical properties (density, viscosity, melt-solid dihedral angle) change as they migrate? Where will they migrate? Their migration paths will have major implications for figuring out the water budget of the Earth - if they are trapped close to or within the slab, hydrating the whole of the mantle transition zone will be much less likely.

There is another reason to be interested in the compositions of hydrous melts. The thermal stability of dense hydrous magnesium silicate (DHMS) phases and the partitioning of water between silicate melt and nominally anhydrous minerals (NAMS) control water transport into the deep mantle. At shallow depths, the fluids released during dehydration of these phases contain only a few percent of dissolved solutes. As such, it is reasonable to simplify thermodynamic calculations by approximating the fluid product of dehydration as pure water. As pressure increases, several processes ensure that this is no longer the case. Firstly, the second critical endpoint of peridotitic and eclogitic systems occur at <10 GPa, after which there is no longer a distinction between melt and fluid, and melt compositions smoothly migrate towards lower water contents with increasing temperature. Secondly, the thermal stability of DHMS phases increases. As a result, the final breakdown of DHMS phases at mantle transition zone pressures occurs when MgO-SiO$_2$-H$_2$O melts have H$_2$O activities significantly lower than one, which will also strongly influence the water content of NAMS. Accurate thermodynamic estimates of water storage capacities in mantle rocks thus require models for both the silicate phases and the coexisting melts.

% There has been some controversy over whether slabs transport water beyond about 300 km depth. Certainly most of the ca. 7 km thick mafic crust is probably too warm to sustain hydrous minerals. However, hydrated ultramafics can potentially carry water into the lower mantle. Supporting evidence for an ultradeep part of the water cycle exists in the form of hydrous ringwoodite in diamond, which apparently formed under cold conditions.

% The compositions of melts formed by the breakdown of hydrous phases is important for a full understanding of the deep water system. The thermodynamics of melting in water-rich systems cannot be determined without good estimates of melt compositions. Furthermore, the thermal stability of hydrous phases cannot be predicted from their properties without also having an activity-composition model for the melt phase. More generally, the density of hydrous melts will be highly composition dependent, impacting on their ability to rise through the mantle. Finally, reactions between migrating melts and the surrounding solid will be controlled by the \emph{P-T-X} dependence on melt composition.

Constraining melt compositions remains a major challenge in high-pressure experimental petrology. The primary problem is that hydrous melts are unquenchable, which means that their compositions must be estimated indirectly. Compositions of partial hydrous melts have been estimated using mass balance, diamond traps \citep{MSUP2007}, and the Mg:Si ratio has been estimated by defocusing the beam of an electron probe microanalyser (EPMA) \citep[e.g.][]{YII2004}. Unfortunately, all three of these techniques are accompanied by significant uncertainties, especially in terms of the water content. An alternative technique is to constrain the liquidus temperature at a given composition, thus circumventing the need to estimate water concentrations in the melt after the experiment (when much of the water has been lost). The problem with this technique is that at ambient pressures, hydrous phases are limited in their maximum water content. In the MgO-SiO$_2$ system, the lack of a thermally stable stoichiometric SiO$_2$.H$_2$O compound means that brucite has been used as the water source. This limits water concentrations in model peridotitic systems (Mg$\sim$1.4 Si) to $\sim$18 wt\% H$_2$O. Thus, studies either extrapolate high temperature liquidus curves to high water contents or add water in liquid form. Extrapolation is prone to large errors, and accurately adding liquid water to the small volumes of high pressure capsules is essentially impossible.

@@

A number of models have been presented to investigate the behaviour of concentrated hydrous melts. \cite{SS1985} made the assumption that H$_2$O, O$_{br}^{2-}$ and OH$_{tm}^-$ mix ideally, with a parameter $r$ representing the number of oxygen atoms in each formula unit which are available for protonation. For example, an Mg$_2$SiO$_4$ melt with $r=4$ and $K=\infty$ represents a continuum between an anhydrous melt and one made of Mg(OH)$_2$ and Si(OH)$_4$ monomers. No distinction is made between isolated oligomers and partial depolymerisation of a silicate network; in other words, the energy required to form a terminal OH group is independent of local environment. Fixing $r$ places an implicit constraint on the maximum value of $K$ in water-rich compositions. For example, if $r=1$ in a hydrated MgO melt, then water contents exceeding 50 mol\% require that $K<\infty$. In practise, the equilibrium constant $K$ has been shown to be dependent on temperature and composition. \cite{SK1995} obtained a good fit to isochemical data with an expression of the form
\begin{equation}
\Delta G^{\circ} = a + b\,T
\end{equation}

@@

In these models, equilibrium melt compositions are calculated by calculating the activity of a component $x$ in the melt which has the composition of the anhydrous liquidus phase $X$ from the Gibbs energy of melting of $X$ at the temperature of interest:
\begin{equation}
RT \ln a_{x,L}^X = G_{X_l} - G_{X_{s}} = - \int_T^{T_{melt}} S_{X_{l}}-S_{X_s} \, dT
\end{equation}
where $T_{melt}$ is the temperature of melting of the liquidus phase under anhydrous conditions.

In this study, we use a modified version of the molecular dynamics-derived anhydrous melt model of \cite{DKS2013} using melt endmembers MgO and SiO$_2$. For MgO melt, only the reference Helmholtz Free energy needed to be adjusted. For the SiO$_2$ melt, the bulk modulus of the SiO$_2$ endmember was increased to satisfy the experimental constraint that coexisting melt and coesite have similar volumes at 13--14 GPa. As a result, SiO$_2$ volumes at lower mantle conditions will be significantly higher than predicted, but this detail is not important for the present study. The volume at 1 bar is fit to experimental data. The entropy at reference conditions is adjusted by the value corresponding to the difference between the experimentally-derived and MD-calculated stishovite values. Remarkably, the value of the entropy derived in this way is essentially identical to that calculated for the experimental data of \cite{} near the coesite-stishovite-melt triple point. As such, we are confident that this value is accurate. It should be noted that the recent melting curve proposed by \cite{} (using a Simon's equation) predicts a much higher entropy of melting at 14 GPa. This is a result of downweighting the data in the study of \cite{}, which indicates a controversially high dT/dP. Nevertheless, this study is not inconsistent with higher pressure studies \citep{} if the gradient of the melting curve shallows between 30 and 70 GPa, which may be the consequence of changing Si-coordination and resultant densification in the melt.

A subregular solution model was adopted to describe the free energy between the MgO and SiO$_2$ melts. The two interaction parameters W$_{MS}$ and W$_{SM}$ 14 GPa were derived from the melt composition at the forsterite (fo)-HP clinoenstatite (hen) eutectic \citep{} and melting temperatures at the fo, hen and fo-hen cotectic compositions. The temperature dependence around 14 GPa was calculated by assuming an ideal excess entropy contribution, and then fitting to the known melting curves of forsterite and enstatite. The pressure dependencies were fit to the melting curves. It should be noted that although the excess volumes are similar to the results of \cite{DKS2013}, the excess entropies are some 50\% lower. However, they provide an excellent fit to the forsterite and orthoenstatite melting curves (see Supplementary Materials).

All of the calculations in the work were undertaken with the mineral physics toolkit Burnman \cite{}.

@@

\subsection{Subsystems}
\subsubsection{MgO-H$_2$O}
The dehydration of brucite at high pressure was previously studied by \cite{FIYKFO2005}. They reported that the stability of brucite reaches a maximum at about 9--10 GPa and 1200 $^{\circ}$C, decomposing at 1100--1150 $^{\circ}$C at 13 GPa. We observed large brucite crystals at 1200 $^{\circ}$C and 13 GPa; the temperature difference is probably within experimental uncertainties. In the present study, the composition of the fluid coexisting with brucite and periclase is constrained to be 64--66 mol\% H$_2$O. The MgO content of the fluid increases gradually with increasing pressure, as a result of the extremely high melting temperature of periclase. The extremely high melting point of MgO means that the best experimental estimate of the melting point at high pressure (7 GPa) is derived by extrapolation from iron-bearing (Mg,Fe)O, using the compositions of coexisting melt and solid to constrain the shape of the two-phase region \citep{ZF2008}. Their melting curve is within error of that obtained from early molecular dynamics simulations \citep{CG1994}, which suggest a melting temperature of $\sim$ 5373 K at 13 GPa. In this study, we follow more recent MD simulations \citep{KKS2013, DKS2013} which predict periclase melting at a lower temperature of $\sim$ 4180 K at 13 GPa, similar to that determined by previous ab-initio studies \citep{Alfe2005}, but significantly higher than early experimental estimates \citep[$\sim$ 3100 K][]{ZB1994}.

At 13 GPa, the calculated entropy of melting of periclase is $27.5$ J/K/mol, somewhat higher than the value from previous MD simulations \cite[$\sim$22 J/K/mol][]{CG1994}. This higher value will moderate the differences in melting temperature between the two studies. Using the ideal OH$^-$ mixing model of \cite{SS1985} with no compositional dependence on the formation of OH groups, $G^{\circ}= -75000$ provides a good fit to the data if additional asymmetry is added via a binary subregular model (W$_{MgO-H_2O}$=55000 J/mol, W$_{H_2O-MgO}$=0 J/mol). This asymmetry is required to fit the activity of water at the periclase-brucite-liquid triple point to the solid thermodynamic models of \citep{HP2011} and the equation of state of water formulated by \cite{PS1995} ($0.52$). This model also has the effect of increasing the tendency for melt-fluid immiscibility at the water-rich end of the system. The second critical endpoint of the MgO-H$_2$O is probably around 13 GPa \citep{MSUP2007}, so large positive interaction enthalpies are justified in this region of the phase diagram.

@@

One chamber from an experiment run at 1300 $^{\circ}$C revealed a single very small crystal of enstatite. Another run at 1200 $^{\circ}$C revealed large crystals of hydroxychondrodite in the chamber with the lowest water contents. Small differences in Mg:Si ratio could explain the differences in solid phases between the separate chambers, so it is assumed that the enstatite+forsterite and chondrodite+forsterite cotectics cross the forsterite+H$_2$O binary at high water contents and 1200--1300 $^{\circ}$C. This supposition is supported by \cite{DDFK2005}, who observe a double-crossing of the liquid and wadsleyite-water tie-line at 15 GPa, first at ca. 1100 $^{\circ}$C, and again at 1400 $^{\circ}$C. Very similar Mg:Si ratios were observed by \citep{LSOK2011} at 15 and 18.5 GPa.

@@

\section{Discussion}
\subsection{Water partitioning in the mantle transition zone}
% Olivine, wadsleyite and ringwoodite

\cite{DDFK2005}
\citep{LSOK2011}
\citep{OMY2000}

\subsection{Deep water release from subduction zones}
% Densities
\cite{MF2010}
\cite{BK2003}
\cite{MSK2008}

% Viscosities
% Shear viscosity 
% Density

%So at least in my models there are three quantities that are important for the pattern of melt migration, the permeability, the compaction viscosity and the shear viscosity (if you have elasticity/plasticity, other things come in, too). Everywhere in the equations where you have the permeability, it's divided by the melt viscosity, and the permeability itself is normally assumed to scale with phi^3 (or less common phi^2), and at least I didn't see any parametrisations describing the effect of temperature on the permeability, so I would assume it's small. So for the permeability, you can just use the values you get for your different P-T-X conditions, and see if the overall term permeability/melt_viscosity increases or decreases.

%For the viscosities of the solid you have the influence of temperature on the viscosity (which I think is similar for shear and compaction viscosity) and you have the effect of porosity, eta ~ exp(alpa * phi) with phi~-27 for the shear viscosity (I think this is what Rich Katz uses in his models, but I would have to look it up to be sure), and xi ~ phi0/phi for the compaction viscosity, and you can do the same here, compare how they change for your different P-T-X conditions.
